---
title: "02- Report"
author: "Minnie Jacobs, Kateryna Chorna, Jemma Hodgkins, Isobelle Clemmens, Gracie Belmonte"
date: "2025-10-25"
output:
  html_document:
    toc: true
    theme: united
    number_sections: true
bibliography: references_lit.bib
---

```{r libraries, include=FALSE}
# Install packages if needed
if(!require("ggplot2")) install.packages("ggplot2")
if(!require("dplyr")) install.packages("dplyr")
if(!require("e1071")) install.packages("e1071")
if(!require("ggrepel")) install.packages("ggrepel")
if(!require("kableExtra")) install.packages("kableExtra")
if(!require("plotly")) install.packages("plotly")
if(!require("caret")) install.packages("caret")
if(!require("stringi")) install.packages("stringi")
if(!require("randomForest")) install.packages("randomForest")
if(!require("glmnet")) install.packages("glmnet")
if(!require("kernlab")) install.packages("kernlab")
if(!require("MASS")) install.packages("MASS")
if(!require("Metrics")) install.packages("Metrics")

# We load the libraries we will need throughout the project:
library(ggplot2)
library(dplyr)
library(e1071)
library(ggrepel)
library(kableExtra)
library(plotly)
library(caret)
library(stringi)
library(randomForest)
library(glmnet)
library(kernlab)
library(MASS)
library(Metrics)
```

```{r loading, include=FALSE}
library(dplyr)
data=read.csv("who_data.csv")
who_data=subset(data,select=-c(X,X.1,X.2,X.3,who_ms))
#convert numeric-like columns stored as character to numeric
num_cols= c("pm10_concentration","pm25_concentration","no2_concentration",
              "pm10_tempcov","pm25_tempcov","no2_tempcov","population")
who_data[num_cols] =lapply(who_data[num_cols], as.numeric) # set the data as numeric

who_data$type_of_stations =as.character(who_data$type_of_stations)
who_data$who_region =as.factor(who_data$who_region)

#and finally, we can see Turkey isn't defined properly in the dataset, so we add 
who_data$country_name[who_data$iso3 == "TUR"] <- "Turkey"
```
# Predicting Pm10 concentration levels for 2021, using data from different cities in Europe

# Introduction
Throughout this project, we analyse air pollution data from the World Health Organisation, focusing on PM10 concentrations as these pollutant levels are a clear indicator of air quality.

The aim of our project is to analyse various models that can be used to predict PM10 concentration levels for cities in Europe, as this area provided the most complete data, making our results more interpretable.
Each model is trained on 90% of our data (using set.seed to ensure we train on the same data), then evaluated on a final validation set. 
Since our dataset contains a large amount of missing data, this prediction is particularly useful for estimating PM10 levels in cities with incomplete observations, allowing us to capture pollution trends across a wide range of European areas.
Each team member applied a different modelling strategy to optimize performance using the same performance metric, we will discuss the suitability of different performance measures such as RMSE for continuous predictions.
Although this modelling approach could be extended to other pollutants or regions, we acknowledge that training only on European data limits generalisability to other continents due to differing emission sources and environmental conditions.
 
## Predictive modelling 
Before we delve into the different models we used, we still have an issue with the missingness of our data. Although we have identified that European cities are more easily analysed since they have a sufficient amount of their data, some predictive models won't work with any missing data at all. 
Hence, we have decided to split our challenge into 2 main categories: 

*Predictive modelling with missingness* - here we simply carry out our predictive modelling by using techniques that will still function despite an NA count. 

```{r europe clean data} 
library(dplyr)
europe_data= who_data %>%
  filter(who_region=="4_Eur")%>%
  select(-iso3, -pm10_tempcov, -pm25_tempcov,-who_region,-no2_tempcov, -type_of_stations)%>%
  mutate(log_population = log(population)) # EDA showed a transform was a better predictor of results

```

*Predictive modelling by removing NAs* - We remove the missing data all together, evidently this creates an over smoothing issue, but for the purpose of this task we will still be able to illustrate our results clearly. 

```{r data set NAs removed} 
clean_data_europe= europe_data %>% select(-pm25_concentration) %>%
na.omit()
```

Note: we can see how many rows were removed, there is a key difference between how many NA values were removed and how many rows were removed, as some rows may contain multiple missing values. 

```{r how many rows removed}
library(dplyr)
europe_rows=nrow(europe_data)
rows_removed=europe_rows-nrow(clean_data_europe)
prop= rows_removed/ europe_rows
prop
```
Hence we have removed 71.6% of our data, so we acknowledge that our models might not be as accurate.

## Limiting covariates 

Following from previous EDA we have carried out, we decided to include the following covariates in our model as they demonstrated significance in the prediction of pm10 concentration levels:

- **year** (temporal, numeric, discrete)- this variable is essential in capturing trends in pm10 pollutant levels over time. 

- **no2_concentration** (environmental)- we found this was a key environmental predictor given it is strongly correlated with pm10_concentration, typically demonstrating an association between the pollutant source.

- **latitude/longitude** (spatial)- important indicators of location, which help us to infer regional differences in pm10 pollution patterns. 

- **population/log-population** (demographic, numeric, continuous)-  We include this as an indicator of city size. 
*Due to its large range and skew, we found using a log transform of this improved our results, so this is also included in our model. *

Coverage variables (pm10_tempcov, no2_tempcov) were excluded since they are related to data completeness rather than directly contributing to prediction.
Note: We have filtered by european city so can exclude continents, we include 
-  **country_name** - (categorical), we have filtered by European countries as these had the highest amount of data, making our results more interpretable. While we don't use it directly in our model, we will use it in visualisations of our results.

## Performance metric 

We read and reviewed a paper discussing various performance metrics in regression models, @performancemetric, the paper outlined RMSE Root mean squared error) and MAE (Mean absolute error) as key metrics to evaluate accuracy between models. 

MAE provides an effective measure of average prediction error, however it is less sensitive to outliers due to the fact all errors are weighted equally. This makes it more robust than RMSE, but less realistic in a real world scenario where outliers are likely to be significant to our predictions.

RMSE is less robust than MAE, but penalises larger deviations, making it a better metric for capturing real world generalisability, for example if we chose to extend our models beyond Europe. 

Given our EDA that we carried out, there was a wide range in pollutant concentration for cities in Europe, so some data points could be interpreted as outliers. Because of this, we focused on RMSE as larger errors in high pollution cities are more impactful in our health setting, however we report all 3 to get a better overview of our models, before concluding in '04-Conclusion'.

## Testing and validation
To assess the performance of our models reliably and ensure their robustness, we must first split our data into two subsets: a test set and a training set. 

- The **training set** consists of 90% all data from our cleaned Europe data, we will use this to perform our k-fold CV. 

- Our **validation set** is 10% of the data that we randomly select which we use for out of sample validation later. 

We use the validation set to see how well our model performs generally, which we could use to predict pm10 concentration for other citities. We note that our models may perform well within Europe, but may not generalise as well outside of this continent.

```{r model split}
set.seed(2)
library(caret)
#for models that include missing data 
train_index <- sample(seq_len(nrow(europe_data)), size = 0.9 * nrow(europe_data)) #creating a 90-10 split
train_data <- europe_data[train_index, ]
test_data  <- europe_data[-train_index, ]

#for models that don't include missing data
train_index_clean <- createDataPartition(clean_data_europe$pm10_concentration, p = 0.9, list = FALSE) #we create a 90-10 split
train_data_clean <- clean_data_europe[train_index_clean, ]
test_data_clean <- clean_data_europe[-train_index_clean, ]

#now using cross validation:
train_cv= trainControl(method = "cv",number=10)
```

# Modelling with missingness 

## Model: Random forest

The use of decision trees is both simple and intuitive. They provide a
straight forward, visual way to represent how different variables
influence a certain outcome by splitting data into small groups based on
predictor values. However, single decision trees can be unstable and
sensitive to training data. This problem is referred to as over-fitting,
where a model performs well on known data but poorly on unseen data. And
so minor changes in the dataset can lead to major changes in the
predicted outcomes. Hence this makes them inefficient when generalising
new, unseen data .

Random Forest trees presents a way of overcoming these issues. The idea
is that they combine the predictions of many single decision trees into
one robust and accurate model. Instead of relying on a single tree, a
Random Forest builds multiple trees using different samples of the data
and random subsets of features at each split. The results from all trees
are then averaged (for regression) or voted on (for classification) to
generate the final prediction.

With this in mind, we can confidently say that this model will be highly
appropriate for our aim of predicting PM10 concentrations across
European cities. Data regarding air quality is complex. It can contain
non-linear patterns, interactions between pollutants and regional
differences. A single regression model may struggle to capture these
elements, but Random Forests can adapt to these complexities without
requiring strong assumptions about the data's structure.@ibm

### Random Forest Model
```{r cross val RF}
#USE CROSS VALIDATION
# Define formula 
rf_formula <- pm10_concentration ~ pm25_concentration + no2_concentration +
               population + latitude + longitude
```

### Train the Random Forest Model

This section implemented the Random Forest algorithm, adapted from @statology

-   The model was specified as:

**PM10 concentration** \~ **NO₂ concentration** + **pm25
concentration** + **population**

-   The algorithm automatically handled missing values using the `na.roughfix`
    function that was found from @stack1
-   The default of 500 decision trees was used.
-   When building each tree, the `randomForest` algorithm only considers
    a random subset of the available predictors at each split.
-   This randomness prevents dominant predictors, for example
    `pm25_concentration` from being used in every tree, and thus skewing
    results.
-   The final prediction is the average of all the trees' output.
```{r}
library(randomForest)

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = train_data, # this will need the whole 90% train data 
  na.action = na.roughfix,  # automatically handles missing predictors   
  trControl = train_cv       
)
rf_model
```
The model output shows:

-   **Mean of Squared Residuals = 28.802** - representing the average
    squared difference between predicted and actual PM10 values.

-   **% Variance Explained = 78.50%** - indicating that almost 80% of
    the variance in PM10 levels is explained by NO2 concentration,
    population, longitude and latitude.

-   However, we note that the **model's effectiveness** will instead be assessed using
    predictions on the 10% unseen test data.

###  Visualising our predictions

-   In this section, the fitted Random Forest model is used to **predict
    PM10 concentrations** .
-   This procedure again uses code from @statology.

```{r predict rf}
#making our predictions
# use the fitted Random Forest model to predict PM10 values for 2021 
predictions <- predict(rf_model, newdata = test_data)
# add the predicted PM10 values to the test dataset 
test_data$predicted_pm10 <- predictions
#head(test_data)
```

-   The `predict()` function in R takes the fitted Random Forest model
    object (`rf_model`) and applies it new data (`test_data`).

-   `test_data\$predicted_pm10 \<- predictions` creates a new variable
    within 'test_data' data-frame that stores the model's predicted PM10
    values.

```{r}
#scatter plot of predicted vs actual PM10  
ggplot(test_data, aes( x = pm10_concentration, y = predicted_pm10)) +
  geom_point(colour = "steelblue", size = 3, alpha = 0.7) +
  geom_abline(slope= 1, intercept = 0, colour= "red", linetype = "dashed") +
  labs(
    title = "Predicted vs Actual PM10 concentrations ",
    x = "Actual PM10 Concentration",
    y = "Predicted PM10 Concentration"
  ) + 
  theme_minimal(base_size = 13)
```


#### Comments on the scatter plot

-   The majority of the points cluster closely around the red line (perfect prediction line), suggesting a good alignment of the predicted and observed values. 

-   This indicates that the Random Forest model successfully captures the key relationships between PM10 and the predictor variables. 

-   The tightness of the spread represents a low overall error, however there are a few outliers present for the higher pollution levels. 

-   These high pollution levels are predicted less accurately, likely because such extreme pollution levels are under-represented in the training data. 

-   Overall, this is a good plot and shows a strong predictive performance of the Random Forest model.

###  Evaluate Model Preformance with a metric 

In order to assess the performance of my chosen model submission, the
Random Forest model, our group has chosen to use the **Root Mean Squared
Error (RSME)**.

We have found the RSME, R^2 and MAE using code adapted from @stack2, @datacamp and @stack3 (respectively).

```{r}
RSME <- sqrt(mean((test_data$pm10_concentration - test_data$predicted_pm10)^2, na.rm = TRUE))

cat("Root Mean Squared Error (RMSE):", round(RSME, 3), "\n")

#calculate R^2 

model_r2 <- lm(test_data$predicted_pm10 ~ test_data$pm10_concentration)
R2 <- summary(model_r2)$r.squared
cat("R-squared (R²):", round(R2, 3), "\n")

# Calculate MAE
MAE <- mean(abs(test_data$pm10_concentration - test_data$predicted_pm10), na.rm = TRUE)
cat("Mean Absolute Error (MAE):", round(MAE, 3), "\n")
```

-    The value of **2.826** tells us that, on average, the Random Forest's predicted PM10 concentrations differ from the true observed values by about **2.8$\mu g/m^3$**.

-    Given that the typical range of PM10 concentrations across Europe regions is often between 10-40$\mu g/m^3$, a value of 2.83 is a relatively small error. 

-    This results supports the visual evidence from the scatter plot, confirming that the model's performance is strong and reliable.

-    Additionally, an $R^{2}$ value of **0.883** tells us that the model explains 88.3% of the variation in PM10 concentrations.
-    Also an MAE of *1.78* means that, in a similar sense, the model’s predicted PM10 concentrations differ from the actual observed values by about 1.78 units, again indicating a relatively small average prediction error and good overall accuracy.

-    This again suggests a strong fit between predicted and actual values.@datacamp

### Analyse Feature Importance 

A useful strength of the Random Forest model is that they can provide a
direct measure of **feature Importance**, in other words, quantifying
how much each variable contributed to the models prediction.

This is calculated using how much each variable reduces the model's
prediction error across all trees in the forest. We used code adapted
from @stackEx.

```{r}
# extract feature importance from the Random Forest Model 

importance_df <- importance(rf_model) %>%
  data.frame() %>%
  mutate(Feature = rownames(.))


#Display
colnames(importance_df)

#Plot Feature Importance 
library(ggplot2)

ggplot(importance_df,aes(x = reorder(Feature, IncNodePurity), y = IncNodePurity)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  coord_flip()+
  labs(
    title = "Feature Importance in Random Forest Mode", 
    x = "Predictor Variable",
    y = "Importance (Increase in MSE)"
  ) +
  theme_minimal(base_size = 13)
```


#### Discussion of the graph

-   The x-axis represents the importance score `IncNodePurity`, which is
    the amount each variable improves the model's ability to make
    accurate splits across all trees in the Random Forest.
-   The y-axis lists the predictors: `longitude`, `latitude`,
    `pm25_concentration`, `no2_concentration` and `population`.

#### Interpretation

-   `Longitude` was the most influential variable, the model relies
    heavily on a location's east-west position to predict PM10. This suggests the presence of clear geographic patterns in air pollution across Europe, possibly linked to regional industrial activity or atmospheric conditions. 
-   `population` has the lowest importance score, implying that population density plays a minor role once environmental and geographical variables are considered.

### Conclusion

In conclusion, by combining multiple decision trees to capture
non-linear patterns between air pollution, geography and population, the
Random Forest model provided an effective approach to predicting PM10
levels across Europe.


With an RSME of **2.8$\mu g/m^3$**, the model achieved a strong level of predictive accuracy, confirming that this algorithm gives a good generalisation to unseen air quality data. 
Overall, we can confidently say that the Random Forest proved to be a robust and realistic modelling approach for our air quality regression problem. 



## Model: XGboost
This model can be found in '03 Report_Python', we discuss our findings in the conclusion

# Modelling without missingness 

## Model: Linear regression (with regularisation)

In this section, we use linear regression to model pm10 concentration across European cities using several explanatory factors such as no2 concentration, population, and geographic features such as longitude and latitude. 
We then extend our regression exploration to Ridge, LASSO, and Elastic Net to address potential multicollinearity and improve predictive performance. We evaluate our model's performance using k-fold cross-validation to train our model, which uses out of sample validation on data from our dataset to assess the generalisation of our model.

From previous EDA that we carried out, we found that using the 'openair' package (https://rdocumentation.org/packages/openair/versions/2.19.0) showed a negative linear trend between pm10 concentration and time, suggesting that linear regression is an appropriate modelling approach. 

Linear regression cannot work if we have missing values, so for this we use our clean_europe_data.

### Model equation
Since we have identified our significant covariates, we now consider transforms and non-linear relationships to create additional variables that can capture more information within the model: 
*no2_concentration: this doesn't have a strictly linear relationship so we include non-linear terms in our model to capture this.*

*longitude and latitude: These are strongly related so we consider their interaction term as a key variable in our linear model. *

So we can give our model using the following equation: 

$$
\text{PM}_{10,i} = \beta_0 + \beta_1 \cdot \text{Year}_i+\beta_2 \cdot \text{Year}_i^2 +\beta_3 \cdot\text{No2}_i+\beta_4 \cdot\text{No2}_i^2+\beta_5 \cdot\text{No2}_i^3+\beta_6 \cdot \text{population}+\beta_7 \cdot \text{log_population}+\beta_8 \cdot \text{longitude}+\beta_9 \cdot \text{latitude}+\beta_{10} \cdot \text{longitude}* \text{latitude}_i+ \varepsilon_i,
$$

where $\varepsilon_i \sim N(0,\sigma^2)$.
```{r model eq}
#using our formula aforementioned:
lm_formula_pm10=pm10_concentration ~ poly(year, 2)+poly(no2_concentration,3) +population+log_population+ longitude+latitude+ latitude*longitude
```

### Test/train

Using the test/train split we outlined earlier, we use our test_data_clean and carry out cross validation on this data set:

```{r cv}
library(caret) #We use the package 'caret' that we saw in the workshop 2.3. 
#now using cv: (we use 10 folds, as we don't have too much data, but it reduces bias)
modelcv=train(lm_formula_pm10, data = train_data_clean, method = "lm" ,trControl=train_cv)

summary(modelcv$finalModel)
```

The estimated coefficients can be interpreted as follows:

-   **Intercept**: $\hat{\beta}_0 = 40.91$\
    This represents the expected PM10 concentration for Europe when all other variables are zero. Since this situation is not realistic, the intercept is not of direct interest.
    
-   **year**: $\hat{\beta}_1 = -64.95$ , $\hat{\beta}_2 = -10.31$, we notice both coefficients of year are negative , suggesting a decrease in pm10 concentrations over time. In the linear term, we have that by holding other variables constant, by increasing the year by 1, we expect a decrease of 64.95 per unit of PM10 concentration. As we saw from our EDA, PM10 concentration appeared to have a negative correlation with year, so this supports our negative relationship (which we could interpret as being perhap due to policy changes). This variable is very statistically significant.
The quadratic term still indicated a decline over time, however this was smaller (10.31), suggesting the rate of decline has decreased in more recent years. This variable did not appear to be statistically significant, but still had a very low p-value.

-   **no2_concentration**: $\hat{\beta}_3 =138.5,\hat{\beta}_4 =163.7,\hat{\beta}_5 =216.6$. 
All of these coefficients were positive and statistically significant, demonstrating a strong non-linear relationship between pollutant concentrations. 
For the linear term, we see that for every increase by 1 unit of NO2, pm10 is expected to increase by 138.5 units, assuming the other factors remain unchanged. Similarly for the quadratic and cubic terms, we see a large increase in PM10 is expected from every unit increase in NO2 concentrations. This could be due to the fact NO2 and PM10 may have the same pollutant sources (e.g traffic). 
    
-   **population**: $\hat{\beta}_6 =-2.935 \cdot 10^{-8} $. This is negative but contributes to a very small change, we also see a very high p-value suggesting population is not statistically significant in the prediction of pm10 concentration.

-   **log_population**: $\hat{\beta}_7 =0.385$ \ This is statistically significant to prediction of pm10 concentration. Holding other variables constant, additional increase in the transformed population by 1 unit is associated with an increase of about $0.385$ units in pm10 concentration. This can be interpreted as more densely populated cities tend to have higher levels of PM10 concentrations, which we would expect.
    
-   **Longitude**: $\hat{\beta}_8 =-0.0856$ This small negative value is statistically significant to our model. Interpretation is more difficult spatially, but a negative longitude variable suggests eastern European cities contribute more to PM10 concentrations.
  
-   **Latitude**: $\hat{\beta}_9=-5.61$ This is more statistically significant than longitude and shows that PM10 concentrations decrease as latitude increases, this corresponds to northern cities experiencing less PM10 concentration. 
   
-   **Longitude X Latitude** $\hat{\beta}_{10}=0.01093$ This is more useful as longitude and latitude are strongly related. This has a very small p-value so we can infer that the interaction term is a statistically significant predictor. This helps us interpret spatially and geographically where pm10 concentrations are expected to be higher/lower.

### Model fit

```{r eval results}
modelcv$results
```

We saw from our model that the cross-validated linear regression achieved an RMSE of 6.621294 on the training set. This means that on average, the model’s predictions of PM10 concentrations deviate from the observed values by about $6.621 \mu g/m^3$, this is a relatively high error, suggesting linear regression struggles to capture more complex relationships within the data. 

From the model summary we saw that most predictors were statistically significant, demonstrating that variables such as NO2 and spatial coordinates contribute to the explanation of PM10 concentrations. The only exception being population, which showed no significant effect. However log population showed to be a statistically significant variable, so by transforming we were able to show population density does contribute to PM10 concentration. 

### Model assumptions

We test our model's fit on the linear regression assumptions:
```{r pred}
par(mfrow = c(2, 2))
plot(modelcv$finalModel)
```

As we can see from the model fit:

*Residuals vs fitted :* The residuals are fairly randomly scattered, but with a curved, quadratic distribution suggesting our assumption for linearity doesn't entirely hold. 

*Q-Q residuals:* The residuals mostly lie in a diagonal line apart from at the tails, where we see extreme values that are not explained by the model, removing outliers could improve the performance of our model.

*Scale-location:* The residuals are randomly scattered around the line, but with a curvature heavily dropping around 15 on the fitted values, the constant variance assumption doesn't fully hold.

*Residuals vs leverage:* We see most of our data is clustered around the left hand side of the graph, suggesting we have a left skew. There appear to be a few points of high leverage. 

Hence, we haven't fitted the optimal model since our assumptions have been violated. Because of this, we expect other models to perform better in our 'artificial competition'.

### Extending our linear regression by regularising

After carrying out a literature review on how to improve our model, we came across a paper on using  @CHEN2019104934, where we build upon the ideas of linear regression by regularising, this means we penalise larger coefficients in order to control the variance and reduce the effects of multicollinearity. 

Where $\lambda$ controls the shrinkage, and $\alpha$ indicates the method. 
There are a variety of ways of doing this, we focus on the following methods: 

- Ridge: $\alpha=0$, Reduces the size of all of the coefficients, but keeps them in the model. We have correlation between some predictors so this method should smooth out any instabilities this causes.

- Lasso: $\alpha=1$, This method is similar to Ridge, but can shrink some coefficients all the way to zero, automatically removing less important variables. 

- Elastic (combines Ridge and Lasso): It keeps Ridge’s stability and LASSO’s ability to select variables, balancing the two through a mixing parameter 
$\alpha$.

```{r ridge, echo=FALSE}
set.seed(21)
#first we try ridge regression
ridge_model <- train(lm_formula_pm10,data = train_data_clean,method = "glmnet",
  trControl = train_cv,
  tuneGrid = expand.grid(alpha = 0,lambda = seq(0.0001, 1, length = 50)))

#now LASSO:
lasso_model <- train(lm_formula_pm10,data = train_data_clean,
  method = "glmnet",trControl = train_cv,
  tuneGrid = expand.grid(alpha = 1,lambda = seq(0.0001, 1, length = 50)))

#and elastic net:
elastic_model <- train(lm_formula_pm10,data = train_data_clean,
method = "glmnet",trControl = train_cv,
  tuneGrid = expand.grid(alpha = seq(0, 1, length = 10),lambda = seq(0.0001, 1, length = 50)))

#so now comparing:
comparison <- data.frame(
  Model = c("Linear", "Ridge", "LASSO", "Elastic Net"),
  RMSE = c(min(modelcv$results$RMSE),min(ridge_model$results$RMSE),min(lasso_model$results$RMSE),min(elastic_model$results$RMSE)),
  R2 = c( max(modelcv$results$Rsquared),max(ridge_model$results$Rsquared),max(lasso_model$results$Rsquared),max(elastic_model$results$Rsquared)
  ),
  MAE = c(min(modelcv$results$MAE),
    min(ridge_model$results$MAE),
    min(lasso_model$results$MAE),
    min(elastic_model$results$MAE)))
  
comparison
```
After running the code for our Ridge, LASSO and Elastic Net, we obtained the above table of comparisons. 

Evidently, the values we can see in our table are all are fairly similar, so evaluating the performance of each model is more nuanced than a clearly defined optimal model.

We see that the LASSO method has the lowest RMSE (Root Mean Squared Error) of 6.608980, meaning it produces the smallest average prediction error on the unseen data, making it potentially more accurate.

Regarding the R^2 value, we can see the Linear model has the highest value, suggesting that when we tested our model on our unseen data, it explained 43.87% of the variation in pm10 concentrations across the European cities. 

Meanwhile the method with the smallest MAE (Mean Absolute Error) is the Elastic Net method. Meaning we can deduce that Elastic Net deviates the least from our true PM10 concentrations.

Since we have chosen to focus on RMSE, we conclude the LASSO method performed the best sa it minimised the prediction error. Most of the differences between our methods are only marginal, but we note different methods perform better with differnt metrics. So one single metric may b=not be representative of all method's performance. 
### Model visualisation:

Now using our model we can use our out of sample data to predict pm10 concentration: 
```{r predictions}
# Predict PM10 for all observations in the dataset
test_data_clean$predicted_pm10 <- predict(lasso_model, newdata = test_data_clean)
#now our results:
results=data.frame(
  actual_pm10 = test_data_clean$pm10_concentration,
  predicted_pm10 = test_data_clean$predicted_pm10,
  country = test_data_clean$country_name)
```


We visualise the relationship between actual pm10 values and predicted pm10 values for European cities, we have coloured the cities by their corresponding countries: 

```{r visual}
library(ggplot2)
ggplot(results, aes(
  x = actual_pm10,y = predicted_pm10,color = country)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "black") +geom_smooth(method = "lm", se = FALSE, color = "red")+
  labs(title = "Actual vs Predicted PM10 concentrations for cities in Europe", 
    x = "Actual PM10 concentration",y = "Predicted PM10 concentration",
    color = "Country"
  ) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "right",plot.title = element_text(face = "bold"))
```

In our figure we have included the 1:1 reference line (perfect prediction line) in black, and our regression line in red.

We can see from our figure that most of our data is clustered around the lower left of our data set, suggesting many of our pm10 pollution levels are lower in value. There are many points evenly distributed around the 1:1 reference line, suggesting a fairly good model fit, with the exception of a few outliers in cities located in France, Israel, Ireland and Italy. These may be due to political or meteorological conditions that we haven't considered in our model. We could consider removing these city outlier points, as it would increase the model performance, however this would also introduce a bias to our model. We also note these outlier points are for higher pollution levels, so removing these would lead to a bias towards lower PM10 concentrations.

Overall our model appears to provide a reasonably good prediction for a variety of cities in Europe, however we identify a few outliers which have slightly skewed the prediction accuracy of our model for higher level pm10 concentrations. 


### Conclusions for Linear regression

To conclude our section on linear regression, we found our initial cross validated linear regression model wasn't very accurate, only accounting for 41% of our data. To resolve this we regularised our linear regression model by using Ridge, LASSO and Elastic net. Although these didn't provide too much improvement (due to the fact there was less multicolinearity between variables than we anticipated), the LASSO method appeared to provide the best results to optimise our linear regresion model. 

We also used transforms of variables, such as log population and polynomial terms for both the no2 concentration and year variables. This helped to capture the non-linearity aspects of our model. However we did find that our model assumptions were violated for the linearity assumption, suggesting further transforms or more explanatory variables should be included to improve our model.

Significant predictors we found were year (which is to be expected given the negative correlation of our pm10 concentrations over time), no2 concentrations spatial locations and log population. 

To amend this and further our analysis, we would need to consider additional explanatory variables, these could include: 

1. Weather conditions - temperature, wind speed, humidity. 
2. Industrial emissions. 
3. Traffic density. 

Including these variables could help to address the assumption violations we found in our model, potentially increasing the explanatory power of the model and providing us with a better understanding of the factors influencing the pm10 concentration levels for cities in Europe.



## Model: Robust regression

We have seen in section Linear Regression how to address multicollinearlity and overfitting with the ridge, lasso and elastics net regression. Now we will look at robust regression models, these models perform more reliably when the data contains outliers. This can be particularly useful in environmental data since there can be anomalies in the data due to short term events such as fires or storms or faulty equipment. However, as our data is the average pm10 concentration from the given year in a given city anomalies may be less likely since we already have an average not the individual values from throughout the year.


In our above working we have been using the ordinary least squares residual which aims to minimise the function $\sum_i (y_i - x_i^T \beta )^2$. As we can see from this equation residuals get quadratic penalties and hence very large residual can dominate the fit. 

For the next section we found a source @appliedstat that explained robust and resistant regression and the packages required in R.


Robust regression minimises a weighted loss function so large residuals have less influence. 


For a given density $f$ we define the loss function as $\rho = - \log f$


So the MLE estimator solves $\min_ \mu \sum_i - \log f (y_i - \mu) = \min _\mu \sum_i \rho (y_i - \mu)$, we set $\psi$ to be the derivative of $\rho$ so $\psi = \rho '$. To find the minimum we set the derivative of the function to zero and hence we get $\sum_i \psi (y_i - \hat{\mu}) =0$. We can also write this in the weighted form with $\sum_i w_i(y_i-\hat{\mu})=0$ where $w_i =\psi (y_i-\hat{\mu})/(y_i-\hat{\mu})$ is the weight of each iteration in the algorithm.


In particular we will look at three robust models which each define $\psi$ differently:


Huber: $\chi (x) = \psi (x)^2 = \min (|x|,c)^2$ where $c=1.5$ is the default.

Tukey : $\psi(t) = t[1-(\frac{t}{R})^2] _+^2$ where the value $R= 4.685$ gives a $95%$ efficiency at the normal.

Hampel: is a piecewise function for example $\psi(x) = \text{sgn} (x) \begin{cases} |x| & \ 0<|x| <a \\ a & a<|x| <b \\ a(c-|x|)/(c-b) & \ b<|x|<c \\ 0 & c<|x| \end{cases}$


The main difference between these models is how quickly they downweigh the outliers, Tukey is the most resistant as it completely rejects very large residuals, Hampel takes a middle ground moderately downweighing and only removing very extreme values and Huber is the least aggressive down weighing residuals but never fully removing them.

We will continue using the same equation for our model as in the Linear Regression section.

```{r}
lm_formula_pm10=pm10_concentration ~ poly(year, 2)+poly(no2_concentration,3) +population+log_population+ longitude+latitude+ latitude*longitude
```

Applying the robust regression models described above requires the use of the rlm() function from the MASS library, where we can specify the robustness function by choosing $\psi$ as described in @appliedstat. However we have previously used the train() function from the caret library in order to conduct cross validation on the model however this doesnt support the rlm() function needed to conduct robust regression. Consequently we have to perform the cross validation manually. 

From @statlearn we define k-fold Cross Validation as $CV_{(k)} = \frac{1}{k} \sum_{i=1}^k \text{MSE}_i$, in K fold cross validation we divide the set of observations into k groups of equal size, the first fold is treated as validation set then the model is fit on the remaining k-1 folds. Mean squared error MSE is computed for each fold and an average of this is computed. In order to find the performance metrics we are using we must compute them from the equations:

$RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^n (y_i-\hat{y}_i)^2)} = \sqrt{MSE}$

$MAE = \frac{1}{n}  \sum_{i=1}^n |y_i-\hat{y}_i|$

$R^2 = 1 - \frac{SS_{res}}{SS_{tot}} = 1- \frac{\sum_{i=1}^n (y_i-\hat{y}_i)^2}{\sum_{i=1}^n (y_i-\bar{y})^2}$

where $y_i$ is the actual value, $\hat{y}_i$ is the predicted value from the model used and $\bar{y}$ is the mean of the actual values. We get get the formula:


```{r}
data <- train_data_clean  # clean training dataset
lm_formula_pm10 <- pm10_concentration ~ poly(year, 2) + poly(no2_concentration, 3) +
                 population + log_population + longitude + latitude + latitude*longitude

set.seed(42)
k <- 10
folds <- sample(1:k, nrow(data), replace = TRUE)
cv_rlm <- function(formula, data, psi_function, k_folds) {
  rmse <- rep(NA, k_folds)
  r2   <- rep(NA, k_folds)
  mae  <- rep(NA, k_folds)
  
  for (i in 1:k_folds) {
    train_fold <- data[folds != i, ]
    test_fold  <- data[folds == i, ]
    
    model <- rlm(formula, data = train_fold, psi = psi_function)
    
    pred <- predict(model, newdata = test_fold)
    actual <- test_fold$pm10_concentration
    
    rmse[i] <- sqrt(mean((actual - pred)^2))
    mae[i]  <- mean(abs(actual - pred))
    
    # R²: 1 - SS_res / SS_tot
    ss_res <- sum((actual - pred)^2)
    ss_tot <- sum((actual - mean(actual))^2)
    r2[i]  <- 1 - ss_res/ss_tot}
  
  data.frame(
    Model = deparse(substitute(psi_function)),
    RMSE = mean(rmse),
    MAE  = mean(mae),
    R2   = mean(r2))}
```


### Resistant Regression

In addition to robust regression, resistant regression also performs more reliably when outliers are present in the dataset. However resistant regression doesn't minimise the ordinary least squares equation, it minimises a different loss function which ignores the extreme points rather than just downweighing them.
In particular we will look at:

Least median of squares, LMS: $\min_b \text{median} |y_i - x_ib|^2$ this minimises the median of squared residuals. It can be very inefficient as it has a converging rate of $\frac{1}{\sqrt[3]{n}}$.

Least trimmed squares regression, LTS: $\min_b \sum_{i=1}^q |y_i -x_ib|^2_{(i)}$ where $q=(n+p+1)/2$, this minimises the sum of the smallest q squared residuals.

S-estimation: the solution to $\sum_{i=1}^n \chi (\frac{y_i-x_ib}{c_0s}) = (n-p)\beta$ with small scale s, $\chi$ is usually chosen as $\chi (u) = u^6 -3u^4 + 3u^2,  |u| \leq 1$ and $c_0= 1.548$ and $\beta = 0.5$ are chosen for consistency at the normal distribution of errors.


Comparing these three models, LMS is the most resistance, LTS is moderately resistant and retains a good efficiency and S-estimator is the least aggressive but is more efficient than LTS and LMS.


Similarly as in robust regression we cannot use the lqs() function from the MASS library within train() hence we create a function for CV which uses the lqs() function.


```{r, include=FALSE}
set.seed(42)
k <- 10
folds <- sample(1:k, nrow(data), replace = TRUE)
cv_lqs <- function(formula, data, method_type, k_folds) {
  rmse <- rep(NA, k_folds)
  r2   <- rep(NA, k_folds)
  mae  <- rep(NA, k_folds)

  for (i in 1:k_folds) {
    train_fold <- data[folds != i, ]
    test_fold  <- data[folds == i, ]

    model <- lqs(formula, data = train_fold, method = method_type)
    pred <- predict(model, newdata = test_fold)
    actual <- test_fold$pm10_concentration

    rmse[i] <- sqrt(mean((actual - pred)^2))
    mae[i]  <- mean(abs(actual - pred))

    # R^2
    ss_res <- sum((actual - pred)^2)
    ss_tot <- sum((actual - mean(actual))^2)
    r2[i]  <- 1 - (ss_res/ss_tot)}

  data.frame(
    Model = paste("lqs", method_type, sep="_"),
    RMSE = mean(rmse),
    MAE  = mean(mae),
    R2   = mean(r2))}

```

### Comparing the models

```{r}
set.seed(42)
cv_results <- bind_rows(
  cv_rlm(lm_formula_pm10, data, psi.huber, k),
  cv_rlm(lm_formula_pm10, data, psi.bisquare, k),
  cv_rlm(lm_formula_pm10, data, psi.hampel, k),
  cv_lqs(lm_formula_pm10, data, "lms", k),
  cv_lqs(lm_formula_pm10, data, "lts", k),
  cv_lqs(lm_formula_pm10, data, "S", k)
)

cv_results
```


The RMSE is significantly high for the resistant models particularly LTS, indicating poor performance. In contrast, the robust models performed better especially the huber model. 
Similarly, for the R² metric, where a higher value indicates better fit, the resistant models all gave negative values showing that they perform worse than the mean. The huber model achieved the highest R². For the resistant models we got negative R^2 which was very surprising for the models to perform that badly that hey perform worse than predicting the mean.
For MAE, where lower values are preferred, all robust models outperformed the others, with Huber once again performing best. 

So we conclude that the robust models performed much better than the resistant models. In particular, the huber model performed the best and therefore we select the Huber model to predict the PM10 concentrations. 

### Predicting

We will use the huber model to predict the pm_10 concentration

```{r}
# Predict PM10 for all observations in the dataset
huber_model <- rlm(lm_formula_pm10, data = train_data_clean, psi = psi.huber)
test_data_clean$pred_pm10_huber    <- predict(huber_model, newdata = test_data_clean)

#now our results:
results=data.frame(
  actual_pm10 = test_data_clean$pm10_concentration,
  pred_pm10_huber = test_data_clean$pred_pm10_huber,
  country = test_data_clean$country_name)
```

```{r}
library(ggplot2)
ggplot(results, aes(
  x = actual_pm10,y = pred_pm10_huber)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "black") +geom_smooth(method = NULL , se = FALSE, color = "red")+geom_smooth(method = 'lm' , se = FALSE, color = "blue")+
  labs(title = "Actual vs Predicted PM10 concentrations for cities in Europe", 
    x = "Actual PM10 concentration",y = "Predicted PM10 concentration from huber"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "right",plot.title = element_text(face = "bold"))
```

Similarly to the graph in the Linear Regression section, we have plotted the actual pm10 values vs the predicted pm10 with two smoothing lines, the blue line indicated the linear regression line and the red line  is the automatic smoothing model within ggplot2. At the start we have more clustering of points near to the perfect prediction line indicating these points were predicted better. As the actual pm10 value gets higher the predicted values gets much further from the actual value.


### Conclusion 

We have explored how robust regression assigns lower weights to outliers, reducing their influence, while resistant regression methods largely ignore extreme observations altogether. In our dataset, the resistant regression models performed poorly, producing large errors and low R^2 values, indicating weak predictive ability. The robust regression models, on the other hand, performed significantly better, with the Huber model achieving the lowest RMSE and highest R^2. We therefore conclude that the Huber model provides the best balance between sensitivity to data and robustness to outliers. However, its performance was still limited this could be because the covariates we used in the model don't fully explain the variability in pm10, for example there might be other variables that are not in the dataset that could help explain this, for example temperature or industrial activity. In addition, the model we specified might not fully capture the non linear relationship between the predictors and pm10 concentration.


## Model: Support vector machines


Support Vector Machines are supervised learning models used for both classification and regression problems. They work by finding an optimal boundary, known as a hyperplane, that best separates or predicts data points. The observations that lie closest to this boundary are called support vectors, and they are the most influential in defining the model.

There are two main types of Support Vector Machines:\
- Support Vector Classification (SVC) which used when the target variable is categorical\
- Support Vector Regression (SVR) which is used when the target variable is continuous

Since we aim to predict PM₁₀ pollutant concentrations, which are continuous values, we use the Support Vector Regression (SVR) model. This model finds a smooth function that predicts PM₁₀ levels but SVR cannot handle missing values directly, so rows containing NA's must be removed or imputed before training the model.

We will try different kernels in SVR to see which works best for predicting PM₁₀ levels.

1.  Linear Kernel
2.  Polynomial Kernel
3.  RBF (Radial Basis Function) Kernel

We will start with simple kernels and increase complexity to find the best model for PM₁₀ predictions.

### Linear Kernal SVR

In Linear SVR, we assume a linear relationship between the input features $x_i \in \mathbb{R}^p$ and the target variable $y_i \in \mathbb{R}$. The model seeks a hyperplane $f(x) = w^\top x + b$, that predicts $y$ with minimal error by minimizing the norm $|w|^2$ subject to an $\varepsilon$-insensitive loss.

The linear kernel is defined as $K(x_i, x_j) = x_i^\top x_j$. This kernel implies that no nonlinear transformation of the input space is performed, so the SVR estimates a linear function of the inputs.

We use the linear kernel due to its simplicity, serving as a baseline before introducing more complex kernels such as polynomial or RBF. It allows us to directly evaluate how well a linear model can capture the relationship between features and the target variable.

```{r}
# Fit linear SVR on training data
svr_linear_model <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,
  method = "svmLinear",     # Linear SVM regression
  trControl = train_cv,
  preProcess = c("center", "scale")  # scale features
)
svr_linear_model
```
The code fits a linear SVR model using the train function from the caret package. It applies 10-fold cross-validation specified earlier for performance evaluation and standardizes the input features by centering and scaling. This is done so each has a mean of 0 and a standard deviation of 1, ensuring that all inputs contribute equally to the model.

Here the metrics are extracted and printed so we can then compare it against models with different kernels.

```{r}
rmse_linear <- svr_linear_model$results$RMSE
r2_linear   <- svr_linear_model$results$Rsquared
mae_linear  <- svr_linear_model$results$MAE

# Print all metrics
cat("RMSE (Linear SVR):", rmse_linear, "\n")
cat("R² (Linear SVR):", r2_linear, "\n")
cat("MAE (Linear SVR):", mae_linear, "\n")
```
These results show that the linear model explains a moderate amount of the variance in the target variable but still leaves a fair amount of error, suggesting that a linear function might not fully capture the structure in the data. We’ll use this to compare to polynomial and RBF kernels to see if they can model the nonlinear relationships more effectively.

### Polynomial Kernel SVR

In Polynomial Kernel SVR, the model captures nonlinear relationships between the covariates $x_i \in \mathbb{R}^p$ and the target $y_i \in \mathbb{R}$. The polynomial kernel is defined as $K(x_i, x_j) = (\gamma \, x_i^\top x_j + r)^d,$ where $d$ is the polynomial degree, $\gamma$ is a scaling factor, and $r$ is a constant term. This kernel maps the inputs into a higher-dimensional space, allowing the SVR to fit a nonlinear function while solving a linear problem in the transformed space.\
We use the polynomial kernel to evaluate whether modeling nonlinear interactions improves predictive performance compared to the linear kernel.

```{r, include=FALSE}
# Define 3 fold cv trainControl for hyperparameter tuning
tune_cv <- trainControl(
  method = "cv",   # k fold cross validation
  number = 3,      # 3 fold CV for hyperparameter tuning
  search = "grid",  
  verboseIter = FALSE
)
# Fit polynomial SVR model with hyperparameter tuning using 3 fold CV defined abpve
svr_poly_model <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,  
  method = "svmPoly",       # Polynomial SVM regression
  trControl = tune_cv,      # 3 fold CV for hyperparameter tuning
  tuneGrid = expand.grid(
    degree = c(2, 3, 4),     # polynomial degree (d)
    scale  = c(0.01, 0.1, 1),# scaling factor for kernel
    C      = c(0.1, 1)   # cost parameter (C)
  ),
  preProcess = c("center", "scale")  # Scale and center the data
)

```
We use hyperparameter tuning, which is the process of selecting the best values for model parameters that are not learned directly from the training data but control the model’s behaviour and complexity. This is done to ensure the model generalises well to new data, avoiding underfitting or overfitting.

We fit a polynomial SVR model using 3-fold cross-validation for hyperparameter tuning. The training data is split into multiple folds, and the model is trained on all but one fold while the remaining fold is used for validation. This process is repeated for each fold, and the performance metrics are averaged to provide a more reliable estimate of the model’s generalisation ability. The model searches over a grid of hyperparameters: the degree of the polynomial controls the flexibility of the kernel, the scale parameter adjusts the scaling of the inner product in the kernel, and C is the cost parameter that balances model complexity against tolerance for prediction errors. Here, r is not included, which is acceptable because the input features are standardised.
```{r}

# After tuning, fit the final model using 10-fold CV with the best parameters
svr_poly_model_final <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,  
  method = "svmPoly",       # Polynomial SVM regression
  trControl = train_cv,     # 10 fold CV for final model training
  tuneGrid = svr_poly_model$bestTune,  # using the best hyperparameters found in tuning
  preProcess = c("center", "scale")  # scale and center the data
)
svr_poly_model_final
```
We get the result printed for the best hyperparamers.

Here the metrics are extracted and printed so we can then compare it against models with different kernels.
```{r}
rmse_poly <- svr_poly_model_final$results$RMSE
r2_poly   <- svr_poly_model_final$results$Rsquared
mae_poly  <- svr_poly_model_final$results$MAE

# Print all metrics
cat("RMSE (Polynomial SVR):", rmse_poly, "\n")
cat("R² (Polynomial SVR):", r2_poly, "\n")
cat("MAE (Polynomial SVR):", mae_poly, "\n")

```
Our model is already better than the linear one for all the metrics used. 

### RBF (Radial Basis Function) Kernel SVR

Similarly, in the RBF Kernel SVR, the kernel is defined as $K(x_i, x_j) = \exp(-\gamma \|x_i - x_j\|^2),$ where \(\gamma > 0\) controls the smoothness of the resulting function by determining how rapidly the influence of a training example decreases with distance. Unlike the polynomial kernel, which models global interactions through polynomial terms, the RBF kernel captures localized nonlinear relationships by implicitly mapping the inputs into an infinite-dimensional feature space. We use the RBF kernel to evaluate whether modeling such localized nonlinear effects further improves predictive performance compared to both the linear and polynomial kernels.
```{r, include=FALSE}

# Fit RBF SVR with hyperparameter tuning (3-fold CV)
svr_rbf_model <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,  # formula
  data = train_data_clean,   # training data
  method = "svmRadial",      # RBF SVM regression
  trControl = tune_cv,       # 3-fold CV for tuning hyperparameters
    tuneGrid = expand.grid(
    sigma = c(0.01, 0.1, 0.5),  # Range of sigma values to tune
    C  = c(1, 5, 10)         # Range of cost values to tune
),       # Hyperparameter grid
)
```
We use hyperparameter tuning again for this kernel to identify the best combination of parameters that yield the highest predictive performance on unseen data. We tune parameters such as the regularization parameter \(C\), the kernel coefficient \(\gamma\), and, for the polynomial kernel, the degree \(d\). In the R implementation, the parameter `sigma` corresponds to the inverse of the kernel width in the theoretical RBF kernel, where \(\gamma = 1 / (2\sigma^2)\). Thus, tuning `sigma` in the model effectively controls the same behavior as tuning \(\gamma\), influencing how tightly the model fits the data. we use cross-validation to prevent overfitting and the optimal parameter set is then used to train the final SVR model.
```{r}
#Final model with best parameters using 10-fold CV
svr_rbf_model_final <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,      # training data
  method = "svmRadial",         # RBF SVM regression
  trControl = train_cv,  # 10-fold CV for final model training
  tuneGrid = svr_rbf_model$bestTune,         # Use the best hyperparameters from tuning
  preProcess = c("center", "scale")  # scale features
)
svr_rbf_model_final 
```
We then train the final RBF SVR model using the optimal hyperparameters obtained from tuning and apply 10-fold cross-validation to  get a more reliable model that works better on unseen data, and we center and scale the predictors for numerical stability.


Here the metrics are extracted and printed so we can then compare it against models with different kernels.
```{r}
rmse_rbf <- svr_rbf_model_final $results$RMSE
r2_rbf   <- svr_rbf_model_final $results$Rsquared
mae_rbf  <- svr_rbf_model_final $results$MAE

# Print all metrics
cat("RMSE (RBF SVR):", rmse_rbf, "\n")
cat("R² (RBF SVR):", r2_rbf, "\n")
cat("MAE (RBF SVR):", mae_rbf, "\n")

```
This is better than polynomial kernel. We will discuss the results of all kernels in the next section.

### Conclusion

#### Evaluating the Kernels
We will now evaluate the performance of each SVR model using metrics such as RMSE, MAE, and \(R^2\). This will allow us to compare the predictive accuracy of the linear, polynomial, and RBF kernels, and determine which kernel provides the best fit for our data.

```{r}
#metrics for each kernel
results <- data.frame(
  Model = c("RBF SVR", "Polynomial SVR", "Linear SVR"),
  RMSE = c(rmse_rbf, rmse_poly, rmse_linear),
  R2 = c(r2_rbf, r2_poly, r2_linear),
  MAE = c(mae_rbf, mae_poly, mae_linear)
)
print(results)
```
We find that the RBF kernel outperforms the linear and polynomial kernels across all metrics, which is expected given its able to capture complex, nonlinear relationships in the data.

The linear SVR performed relatively poorly, suggesting that the relationship between the covariates and the target variable is not purely linear. During hyperparameter tuning, the polynomial SVR favoured a higher polynomial degree, indicating that the model required a more complex functional form to fit the data effectively. However, this increased model complexity came with a significant computational cost, as significantly increased the training times. In contrast, the RBF kernel captured the nonlinear structure of the data more efficiently, providing a smoother regression function and better overall performance while remaining computationally feasible.

#### Testing RBF SVR on Out of Sample Data
We will now use the out- of sample data to see how well the RBF SVR performs on unseen data and evaluate its generalisation ability. 

```{r}
# Predict on Test Data (unseen)
predictions_rbf_test <- predict(svr_rbf_model_final, newdata = test_data_clean)

```
We plot Actual vs Predicted for each city to visually assess the performance of the RBF SVR model. 
```{r}
# Plot actual vs predicted
p_rbf_test <- ggplot(test_data_clean, aes(
  x = pm10_concentration,
  y = predictions_rbf_test,
  text = paste0("City: ", test_data_clean$city, "<br>Country: ", test_data_clean$country_name)
)) +
  geom_point(color = "darkgreen", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Actual PM10",
    y = "Predicted PM10",
    title = "Actual vs Predicted for Out of Sample Data"
  ) +
  theme_minimal()
ggplotly(p_rbf_test, tooltip = "text")

```
A lot of points lie close to the dashed red line, indicating that the model predicts PM10 concentrations well for many cities. However, some points are quite far from the line, meaning that the model struggles to accurately predict PM10 for certain locations.

```{r}
# Calculate test set metrics
rmse_rbf_test <- sqrt(mean((test_data_clean$pm10_concentration - predictions_rbf_test)^2))
mae_rbf_test  <- mean(abs(test_data_clean$pm10_concentration - predictions_rbf_test))
r2_rbf_test   <- 1 - sum((test_data_clean$pm10_concentration - predictions_rbf_test)^2) / 
                     sum((test_data_clean$pm10_concentration - mean(test_data_clean$pm10_concentration))^2)
cat("Test RMSE (RBF SVR):", rmse_rbf_test, "\n")
cat("Test R² (RBF SVR):", r2_rbf_test, "\n")
cat("Test MAE (RBF SVR):", mae_rbf_test, "\n")
```


We now Turn our attention to our '04_Conclusions' file, where we can see how our models compared against one another. 
# References

