---
title: "04_Conclusions"
author: "Minnie"
date: "2025-10-25"
output: html_document
---

```{r libraries, include=FALSE}
# Install packages if needed
if(!require("ggplot2")) install.packages("ggplot2")
if(!require("dplyr")) install.packages("dplyr")
if(!require("kableExtra")) install.packages("kableExtra")

#libraries
library(ggplot2)
library(dplyr)
library(kableExtra)
```
# Performance metric

Throughout our report we recorded the RMSE, the R^2 and the MAE. 
The RMSE is a good indicator of how large the prediction errors are on average, penalizing larger errors more heavily than smaller ones. This makes it particularly useful in our environmental context, where large deviations can indicate critical pollution events.

$$
\mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
$$

Where $y_i$ are the observed values and $\hat{y_i}$ are our predicted values.


The MAE measures the average absolute difference between predictions and actual observations, treating all errors equally regardless of direction or magnitude, this is less relevant to our environmental context.

$$
\text{MAE} = \frac{1}{n} \sum_{i=1}^{n}{| \hat{y_i} - {y}_i | }
$$


The R^2 measures the proportion of variance in the dependent variable explained by the model

$$
\text{R^2} = 1-\frac{SS_{residuals}} {SS_{total}}
$$
Where $SS_{residuals}$ is the sum of squared errors between predictions and actual values
And $SS_{totals}$ is the total variance in the observed data.

We used RMSE as our chosen performance metric as it took into account outliers, which are most relevant in identifying extreme pollution levels.

However, it is worth noting that although we chose RMSE as our chosen metric, there is value in other metrics that aren't captured solely by the RMSE, so we have included all three in our comparison tables, but focus our final discussion on the RMSE of each method. 


## Our results 

We each explored a seperate model to optimize a predictive model to predict PM10 concentration for cities in Europe. We focused our modelling on 2 key areas: 

- *Models with missingness*, our results are displayed below for all performance metrics to 3dp:

```{r w missingness}
model_results_NAs <- data.frame(
  Model = c("Random Forest", "XGBoost"),
  RMSE = c(2.830,4.832),
  MAE = c(1.786,2.684),
  R2 = c(0.883,0.818))

#table 
model_results_NAs %>%
  kbl(caption = "Comparison of performance metrics for models with missingness ") %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover", "condensed"))
```

The Random Forest model provided the best results across metrics, explaining nearly 88% of the variance in PM10 concentration levels and maintaining a low RMSE, demonstrating its suitability for handling complex, nonlinear relationships and partial missing data.

In comparison, XGBoost performed well but showed higher error values. This is likely due to its sequential tree-building process, which can increase sensitivity to missing data and risk of overfitting when the dataset is sparse. The Random Forest’s built-in randomness through bootstrapped sampling and feature bagging, provided us with more stable predictions under these conditions.

- *Models without missingness:*
 
```{r without missingness}
model_results <- data.frame(
  Model = c("Support Vector Regression (RBF)", "Linear regression (LASSO)","Robust Regression(Huber)"),
RMSE = c(4.265,6.609,6.737),
  MAE = c(4.265, 4.726,4.643),  
  R2 = c(0.704, 0.437, 0.413))

model_results %>%
  kbl(caption = "Comparison of performance metrics for models without missingness") %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover", "condensed"))

```

As we can see models that use missingness performed better than those that remove it, this is to be expected as they have more information.
Among our regression-based approaches, the Support Vector Regression (RBF kernel) achieved the strongest performance across all metrics, capturing nonlinear spatial and environmental patterns more effectively than linear models.  
Although we carried out LASSO regularisation on linear regression to improve generalisation, there wasn't any particular improvement from the standard linear model. 
 The Robust regression approach handled outliers effectively but could not fully capture nonlinearity.

# WHO data context

The best-performing models, particularly Random Forest, could be applied to estimate missing PM10 concentrations in European cities, helping identify areas where monitoring is limited but pollution risk remains high.
This is particularly relevant if this data is used for insights into potential risks these pollutants pose; informative conclusions about pollutant risks can be made even in less developed regions even when data collection is rare.


# Conclusions

Hence, we can conclude that the Random Forest achieved the lowest RMSE (**2.83$\mu g/m^3$**), clearly out performing XGBoost (**4.83$\mu g/m^3$**), SVM (**4.27$\mu g/m^3$**), LASSO Regression (**6.61$\mu g/m^3$**) and the Robust Huber model (**6.74$\mu g/m^3$**). This winning performance is due to number of reasons. 

-    Unlike the other models (excluding XGBoost), Random Forest could train on data containing NA's using `na.roughfix`. This means the model initially had access to more complete information, allowing the model to learn broader patterns in the data, hence creating more accurate predictions. 

-    Our PM10 levels depend on interacting environmental, spatial and demographic predictors. The geographical variables like `latitude` and `longitude` interact in complex, non-linear ways. Hence these patterns mean that linear and SVM models (with linear kernels) struggle to fit the data accurately. However, as we have seen, the Random Forest can model these interactions effectively, thus proving a more successful model than the others. 

-  Random Forest outperforms XGBoost, although many papers suggeted XGBoost often performs better, we theorise that it is due to the fact its bootstrapped sampling and feature bagging made it less sensitive to noise and missingness in our relatively small dataset.
Each tree in a Random Forest is built using a random subset of both observations and predictor variables, meaning no two trees will use the same data, reducing correlation. XGBoost, in contrast, builds trees sequentially to correct previous errors. This can make it more prone to overfitting when the dataset contains a high proportion of missing values or noisy features, as was the case in our WHO dataset.


-  Overall, the Random Forests' ability to handle missing values and non-linear patterns along with it's flexibility, randomness and robustness made it the most reliable and realistic model for forecasting PM10 concentrations across Europe.
