---
title: "04_Conclusions"
author: "Minnie"
date: "2025-10-25"
output: html_document
---
```{r libraries, include=FALSE}
# Install packages if needed
if(!require("ggplot2")) install.packages("ggplot2")
if(!require("dplyr")) install.packages("dplyr")
if(!require("kableExtra")) install.packages("kableExtra")

#libraries
library(ggplot2)
library(dplyr)
library(kableExtra)
```

# Performance metric
Throughout our report we recorded the RMSE, the R^2 and the MAE. 
The R^2 is a good indicator of ... 

$$
\text{RMSE} = \sqrt{ \frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 }
$$

Where y_i is the...
\hat{y_i} are our predicted ... 

The MAE ... 

$$
\text{RMSE} = \sqrt{ \frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 }
$$


However, it is worth noting that although we chose RMSE as our chosen metric, there is value in other metrics that aren't captured solely by the RMSE. 

## Our results 
We each explored a seperate model to optimize a predictive model to predict PM10 concentration for cities in Europe. Our final modelling findings can be found in the table below: 

```{r}
model_results <- data.frame(
  Model = c("Random Forest", "XGBoost","Huber (Robust Regression)", "Support Vector Regression (RBF)", "Linear regression (LASSO)"),
  RMSE = c(2.83,1, 6.737082,4.265415,6.608980),
  MAE = c(NA,1, 4.642810,4.265415,4.726464),     # Random Forest MAE not provided
  R2 = c(0.883,1, 0.4127671,0.70404,0.4366445)
)


# Display the table nicely
library(kableExtra)
model_results %>%
  kbl(caption = "Comparison of Model Performance Metrics") %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover", "condensed"))


```
As we can see models that use missingness performed better than those that remove it, this is to be expected as they have more information.

The Random Forest model provided the best results across metrics, explaining nearly 88% of the variance in pm10 concentration levels and maintaining a low RMSE, demonstrating its suitability for handling complex, nonlinear relationships and partial missing data. XGBoost also performd very well on our out of sample data , ...

Among our regression based approaches, 



Our Support Vector machine achieved a strong performance across all evaluation metrics (RMSE, MAE, and R²). This confirms that nonlinear, kernel-based approaches can better capture the intricate spatial and environmental patterns underlying PM10 concentrations compared to linear models.

To improve ... 





# Conclusions
Hence, we can conclude ...

