---
title: "Kateryna Chorna Support Vector Machine"
author: "Kateryna Chorna"
date: "2025-10-20"
output:
  html_document:
    toc: true
    theme: united
    number_sections: true
---

# Library requirements

In order to run our code, the following libraries are required:

```{r}

# Install packages if needed
if(!require("ggplot2")) install.packages("ggplot2")
if(!require("dplyr")) install.packages("dplyr")
if(!require("e1071")) install.packages("e1071")
if(!require("ggrepel")) install.packages("ggrepel")
if(!require("kableExtra")) install.packages("kableExtra")
if(!require("plotly")) install.packages("plotly")
if(!require("caret")) install.packages("caret")
if(!require("stringi")) install.packages("stringi")

# We load the libraries we will need throughout the project:
library(ggplot2)
library(dplyr)
library(e1071)
library(ggrepel)
library(kableExtra)
library(plotly)
library(caret)
library(stringi)

```

# Data Cleaning

```{r load_and_preprocess, message=FALSE, warning=FALSE}
# Load packages
library(dplyr)
library(e1071)

# Load the data
data <- read.csv("../who_data.csv")

# Remove irrelevant columns
who_data <- subset(data, select=-c(X,X.1,X.2,X.3,who_ms))

# Replace literal "NA" strings with proper R NA
who_data[who_data == "NA"] <- NA

#We can see Turkey isn't defined properly in the dataset, so we add 
who_data$country_name[who_data$iso3 == "TUR"] <- "Turkey"

# Convert numeric-like columns stored as character to numeric
num_cols <- c("pm10_concentration","pm25_concentration","no2_concentration",
              "pm10_tempcov","pm25_tempcov","no2_tempcov","population")
who_data[num_cols] <- lapply(who_data[num_cols], as.numeric)

# Convert categorical variables to correct type
who_data$type_of_stations <- as.character(who_data$type_of_stations)
who_data$who_region <- as.factor(who_data$who_region)

# Filter for European countries and remove pm25 column
europe_data <- who_data %>%
  filter(who_region == "4_Eur") %>%
  select(-pm25_concentration)

# Remove rows with missing values
clean_data_europe <- na.omit(europe_data)

summary(clean_data_europe)


```

```{r}
set.seed(2)

#for models that don't include missing data
train_index_clean <- createDataPartition(clean_data_europe$pm10_concentration, p = 0.9, list = FALSE) #we create a 90-10 split
train_data_clean <- clean_data_europe[train_index_clean, ]
test_data_clean <- clean_data_europe[-train_index_clean, ]
```

# Support Vector Machine (SVM)

Support Vector Machines are supervised learning models used for both classification and regression problems. They work by finding an optimal boundary, known as a hyperplane, that best separates or predicts data points. The observations that lie closest to this boundary are called support vectors, and they are the most influential in defining the model.

There are two main types of Support Vector Machines:\
- Support Vector Classification (SVC) which used when the target variable is categorical\
- Support Vector Regression (SVR) which is used when the target variable is continuous

## Model used in this project

Since we aim to predict PM₁₀ pollutant concentrations, which are continuous values, we use the Support Vector Regression (SVR) model.\
This model finds a smooth function that predicts PM₁₀ levels but SVR cannot handle missing values directly, so rows containing NA's must be removed or imputed before training the model.

Support Vector Regression (SVR) could be useful for environmental data because:\
- It can deal with non-linear relationships.\
- It is quite robust to noise and outliers.\
- It can use several explanatory variables at once to improve predictions.

Even though SVR has these advantages, we will also try other regression and machine learning models to see which one predicts PM₁₀ levels across European countries best.

We will try different kernels in SVR to see which works best for predicting PM₁₀ levels.

1.  **Linear Kernel**
    -   Simple and fast.
    -   Good if the relationship between features and PM₁₀ is roughly linear.
2.  **Polynomial Kernel**
    -   Can capture curved relationships in the data.
    -   Start with low degrees (e.g., 2 or 3) to avoid overfitting.
3.  **RBF (Radial Basis Function) Kernel**
    -   Can model complex, non-linear patterns.
    -   Default choice if the relationship is unknown.
    -   Gamma parameter may need tuning for best performance.

We will start with simple kernels and increase complexity to find the best model for PM₁₀ predictions.

## Define cross validation

```{r}
# 10-fold cross-validation
train_cv <- trainControl(method = "cv",number = 10, verboseIter = TRUE)
```

## Model Evaluation Metrics

-   **RMSE (Root Mean Squared Error):** Measures typical prediction error in µg/m³, penalising large errors. Lower values are better.\
-   **MAE (Mean Absolute Error):** Average absolute difference between predicted and actual PM₁₀. Less sensitive to outliers than RMSE. Lower is better.\
-   **R² (R-squared):** Proportion of variance in PM₁₀ explained by the model. Ranges from 0 to 1, higher is better.

# Linear Kernal SVR

In Linear SVR, we assume a linear relationship between the input features $x_i \in \mathbb{R}^p$ and the target variable $y_i \in \mathbb{R}$. The model seeks a hyperplane\
$f(x) = w^\top x + b$, that predicts $y$ with minimal error by minimizing the norm $|w|^2$ subject to an $\varepsilon$-insensitive loss.

The linear kernel is defined as $K(x_i, x_j) = x_i^\top x_j$. This kernel implies that no nonlinear transformation of the input space is performed, so the SVR estimates a linear function of the inputs.

We use the linear kernel due to its simplicity, serving as a baseline before introducing more complex kernels such as polynomial or RBF. It allows us to directly evaluate how well a linear model can capture the relationship between features and the target variable.

```{r}
# Fit linear SVR on training data
svr_linear_model <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,
  method = "svmLinear",     # Linear SVM regression
  trControl = train_cv,
  preProcess = c("center", "scale")  # scale features
)
svr_linear_model
```

The code fits a linear SVR model using the train function from the caret package. It applies 10-fold cross-validation specified earlier for performance evaluation and standardizes the input features by centering and scaling. \## Model Evaluation

```{r}
rmse_linear <- svr_linear_model$results$RMSE
r2_linear   <- svr_linear_model$results$Rsquared
mae_linear  <- svr_linear_model$results$MAE

# Print all metrics
cat("RMSE (Linear SVR):", rmse_linear, "\n")
cat("R² (Linear SVR):", r2_linear, "\n")
cat("MAE (Linear SVR):", mae_linear, "\n")
```

These results show that the linear model explains a moderate amount of the variance in the target variable but still leaves a fair amount of error, suggesting that a linear decision function might not fully capture the structure in the data. We’ll use this to compare to polynomial and RBF kernels to see if they can model the nonlinear relationships more effectively.

# Polynomial Kernel SVR

```{r}
# Fit polynomial SVR

# Step 1: Define trainControl for hyperparameter tuning (3-fold CV)
tune_cv <- trainControl(
  method = "cv",   # k-fold cross-validation
  number = 3,      # 3-fold CV for hyperparameter tuning
  search = "grid",  # Search over the hyperparameter grid
  verboseIter = TRUE
)


# Step 3: Fit polynomial SVR model with hyperparameter tuning using 3-fold CV
svr_poly_model <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,  # your training dataset
  method = "svmPoly",       # Polynomial SVM regression
  trControl = tune_cv,      # 3-fold CV for hyperparameter tuning
  tuneGrid = expand.grid(
    degree = c(2, 3, 4),     # polynomial degree
    scale  = c(0.01, 0.1, 1),# scaling factor for kernel
    C      = c(0.1, 1)   # cost parameter (C)
  ),
  preProcess = c("center", "scale")  # Optional: scale and center the data
)

```

Training Time?

```{r}

# Step 4: After tuning, fit the final model using 10-fold CV with the best parameters
svr_poly_model_final <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,  # your training dataset
  method = "svmPoly",       # Polynomial SVM regression
  trControl = train_cv,     # 10-fold CV for final model training
  tuneGrid = svr_poly_model$bestTune,  # Use the best hyperparameters found in tuning
  preProcess = c("center", "scale")  # Optional: scale and center the data
)
svr_poly_model_final
```

## Model Evaluation

```{r}

rmse_poly <- svr_poly_model_final$results$RMSE
r2_poly   <- svr_poly_model_final$results$Rsquared
mae_poly  <- svr_poly_model_final$results$MAE

# Print all metrics
cat("RMSE (Polynomial SVR):", rmse_poly, "\n")
cat("R² (Polynomial SVR):", r2_poly, "\n")
cat("MAE (Polynomial SVR):", mae_poly, "\n")

```

Our model is already better than the linear one

# RBF (Radial Basis Function) Kernel SVR

```{r}

# Step 3: Fit RBF SVR with hyperparameter tuning (3-fold CV)
svr_rbf_model <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,  # formula
  data = train_data_clean,   # training data
  method = "svmRadial",      # RBF SVM regression
  trControl = tune_cv,       # 3-fold CV for tuning hyperparameters
    tuneGrid = expand.grid(
    sigma = c(0.01, 0.1, 0.5),  # Range of sigma values to tune
    C  = c(1, 5, 10)         # Range of cost values to tune
),       # Hyperparameter grid
)
```

```{r}
#Final model with best parameters using 10-fold CV
svr_rbf_model_final <- train(
  pm10_concentration ~ year * no2_concentration + population + latitude + longitude,
  data = train_data_clean,      # training data
  method = "svmRadial",         # RBF SVM regression
  trControl = train_cv,  # 10-fold CV for final model training
  tuneGrid = svr_rbf_model$bestTune,         # Use the best hyperparameters from tuning
  preProcess = c("center", "scale")  # scale features
)
svr_rbf_model_final 
```

## Model Evaluation

```{r}
rmse_rbf <- svr_rbf_model_final $results$RMSE
r2_rbf   <- svr_rbf_model_final $results$Rsquared
mae_rbf  <- svr_rbf_model_final $results$MAE

# Print all metrics
cat("RMSE (RBF SVR):", rmse_rbf, "\n")
cat("R² (RBF SVR):", r2_rbf, "\n")
cat("MAE (RBF SVR):", mae_rbf, "\n")

```

This is better than polynomial kernel

# Final SVR Model Evaluation on out of sample data

```{r}
#metrics for each kernel
results <- data.frame(
  Model = c("RBF SVR", "Polynomial SVR", "Linear SVR"),
  RMSE = c(rmse_rbf, rmse_poly, rmse_linear),
  R2 = c(r2_rbf, r2_poly, r2_linear),
  MAE = c(mae_rbf, mae_poly, mae_linear)
)
print(results)
```

we find that thatws the best model

```{r}
# Predict on Test Data
predictions_rbf_test <- predict(svr_rbf_model_final, newdata = test_data)

```

## Plot Actual vs Predicted

```{r}

# Plot actual vs predicted
p_rbf_test <- ggplot(test_data, aes(
  x = pm10_concentration,
  y = predictions_rbf_test,
  text = paste0("City: ", test_data$city, "<br>Country: ", test_data$country_name)
)) +
  geom_point(color = "darkgreen", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Actual PM10",
    y = "Predicted PM10",
    title = ""
  ) +
  theme_minimal()

ggplotly(p_rbf_test, tooltip = "text")

```

## Model Evaluation

```{r}
# Evaluation 
```
