---
title: "RandomForestFinal"
output: html_document
date: "2025-11-03"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load packages 


library(here)
library(dplyr)
library(tidyr) #for pivoting 
library(randomForest)
library(caret)

```

# 1. Introduction

The use of decision trees is both simple and intuitive. They provide a
straight forward, visual way to represent how different variables
influence a certain outcome by splitting data into small groups based on
predictor values. However, single decision trees can be unstable and
sensitive to training data. This problem is referred to as over-fitting,
where a model performs well on known data but poorly on unseen data. And
so minor changes in the dataset can lead to major changes in the
predicted outcomes. Hence this makes them inefficient when generalising
new, unseen data .

Random Forest trees presents a way of overcoming these issues. The idea
is that they combine the predictions of many single decision trees into
one robust and accurate model. Instead of relying on a single tree, a
Random Forest builds multiple trees using different samples of the data
and random subsets of features at each split. The results from all trees
are then averaged (for regression) or voted on (for classification) to
generate the final prediction.

With this in mind, we can confidently say that this model will be highly
appropriate for our aim of predicting PM10 concentrations across
European cities. Data regarding air quality is complex. It can contain
non-linear patterns, interactions between pollutants and regional
differences. A single regression model may struggle to capture these
elements, but Random Forests can adapt to these complexities without
requiring strong assumptions about the data's structure.

reference: [1] "<https://www.ibm.com/think/topics/random-forest>"

# 2. Data Preparation


```{r}

who_data <- read.csv(here("GracieBelmonte","data","who_data.csv"))


europe_data= who_data %>%
  filter(who_region=="4_Eur")%>%
  select(-iso3, -pm10_tempcov, -pm25_tempcov,-who_region,-no2_tempcov, -type_of_stations)%>%
  filter(!is.na(pm10_concentration)) # drop the missing values of the row we are predicting 
  

set.seed(2)

#for models that include missing data 
#split training and test data
train_index <- sample(seq_len(nrow(europe_data)), size = 0.9 * nrow(europe_data)) #creating a 90-10 split
train_data <- europe_data[train_index, ]
test_data  <- europe_data[-train_index, ]


#USE CROSS VALIDATION
# Define formula 

rf_formula <- pm10_concentration ~ pm25_concentration + no2_concentration +
               population + latitude + longitude


# 10 fold cross validation step 
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)


train_cv= trainControl(method = "cv",number=10)

```

The key variables retained were:

-   `pm10_concentration` - the target variable to be predicted.
-   `pm25_concentration`, no2_concentration`, `population`, `longitude` and `latitude` -
    predictor variables.
-   `country_name`, `city` and `year` - identifying variables.

# 3. Train the Random Forest Model

This section implemented the Random Forest algorithm, adapted
from: [2] "<https://www.statology.org/random-forest-in-r/>"

-   The model was specified as:

**PM10 concentration** \~ **NO₂ concentration** + **pm25
concentration** + **population**

`pm10_concentration ~ no2_concentration + pm25_concentraion + population`

-   The algorithm automatically handled missing values using the `na.roughfix`
    function that was found from :
    [3] '<https://stackoverflow.com/questions/8370455/how-to-use-random-forests-in-r-with-missing-values>"
-   The default of 500 decision trees was used.
-   When building each tree, the `randomForest` algorithm only considers
    a random subset of the available predictors at each split.
-   This randomness prevents dominant predictors, for example
    `pm25_concentration` from being used in every tree, and thus skewing
    results.
-   The final prediction is the average of all the trees' output.

```{r}

library(randomForest)

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = train_data, # this will need the whole 90% train data 
  na.action = na.roughfix,  # automatically handles missing predictors   
  trControl = train_control        
)

rf_model

```

The model output shows:

-   **Mean of Squared Residuals = 28.802** - representing the average
    squared difference between predicted and actual PM10 values.

-   **% Variance Explained = 78.50%** - indicating that almost 80% of
    the variance in PM10 levels is explained by NO2 concentration,
    population, longitude and latitude.

-   However, we note that the **model's effectiveness** will instead be assessed using
    predictions on the 10% unseen test data.

# 4. Visualising our predictions

-   In this section, the fitted Random Forest model is used to **predict
    PM10 concentrations** .
-   This procedure again uses code
    from:[2]"<https://www.statology.org/random-forest-in-r/>"

```{r}

#making our predictions

# use the fitted Random Forest model to predict PM10 values for 2021 
predictions <- predict(rf_model, newdata = test_data)

# add the predicted PM10 values to the test dataset 
test_data$predicted_pm10 <- predictions

#head(test_data)



```

-   The `predict()` function in R takes the fitted Random Forest model
    object (`rf_model`) and applies it new data (`test_data`).

-   `test_data\$predicted_pm10 \<- predictions` creates a new variable
    within 'test_data' data-frame that stores the model's predicted PM10
    values.

```{r}

# load ggplot2 for graphs 

library(ggplot2)

#scatter plot of predicted vs actual PM10  
ggplot(test_data, aes( x = pm10_concentration, y = predicted_pm10)) +
  geom_point(colour = "steelblue", size = 3, alpha = 0.7) +
  geom_abline(slope= 1, intercept = 0, colour= "red", linetype = "dashed") +
  labs(
    title = "Predicted vs Actual PM10 concentrations ",
    x = "Actual PM10 Concentration",
    y = "Predicted PM10 Concentration"
  ) + 
  theme_minimal(base_size = 13)



```

### Comments on the scatter plot

-   The majority of the points cluster closely around the red line (perfect prediction line), suggesting a good alignment of the predicted and observed values. 

-   This indicates that the Random Forest model successfully captures the key relationships between PM10 and the predictor variables. 


-   The tightness of the spread represents a low overall error, however there are a few outliers present for the higher pollution levels. 


-   These high pollution levels are predicted less accurately, likely because such extreme pollution levels are under-represented in the training data. 


-   Overall, this is a good plot and shows a strong predictive performance of the Random Forest model.


# 5. Evaluate Model Preformance with a metric 

In order to assess the performance of my chosen model submission, the
Random Forest model, our group has chosen to use the **Root Mean Squared
Error (RSME)**.

We have implemented this using code adapted from:[4]
"<https://stackoverflow.com/questions/60685866/how-do-i-find-out-the-rmse-of-a-random-forest-in-r>"

```{r}


RSME <- sqrt(mean((test_data$pm10_concentration - test_data$predicted_pm10)^2, na.rm = TRUE))

cat("Root Mean Squared Error (RMSE):", round(RSME, 3), "\n")



```

-    The value of **2.826** tells us that, on average, the Random Forest's predicted PM10 concentrations differ from the true observed values by about **2.8$\mu g/m^3$**.

-    Given that the typical range of PM10 concentrations across Europe regions is often between 10-40$\mu g/m^3$, a value of 2.83 is a relatively small error. 

-    This results supports the visual evidence from the scatter plot, confirming that the model's performance is strong and reliable.

# 6. Analyse Feature Importance 

A useful strength of the Random Forest model is that they can provide a
direct measure of **feature Importance**, in other words, quantifying
how much each variable contributed to the models prediction.

This is calculated using how much each variable reduces the model's
prediction error across all trees in the forest. We used code adapted
from
[5]"<https://stats.stackexchange.com/questions/153663/how-to-get-the-most-important-variables-in-random-forests-in-r>")

```{r}

# extract feature importance from the Random Forest Model 

importance_df <- importance(rf_model) %>%
  data.frame() %>%
  mutate(Feature = rownames(.))


#Display
colnames(importance_df)

#Plot Feature Importance 
library(ggplot2)

ggplot(importance_df,aes(x = reorder(Feature, IncNodePurity), y = IncNodePurity)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  coord_flip()+
  labs(
    title = "Feature Importance in Random Forest Mode", 
    x = "Predictor Variable",
    y = "Importance (Increase in MSE)"
  ) +
  theme_minimal(base_size = 13)




```

## Discussion of the graph

-   The x-axis represents the importance score `IncNodePurity`, which is
    the amount each variable improves the model's ability to make
    accurate splits across all trees in the Random Forest.
-   The y-axis lists the predictors: `longitude`, `latitude`,
    `pm25_concentration`, `no2_concentration` and `population`.

### Interpretation

-   `Longitude` was the most influential variable, the model relies
    heavily on a location's east-west position to predict PM10. This suggests the presence of clear geographic patterns in air pollution across Europe, possibly linked to regional industrial activity or atmospheric conditions. 
-   `population` has the lowest importance score, implying that population density plays a minor role once environmental and geographical variables are considered.


# 7. Conclusion

In conclusion, by combining multiple decision trees to capture
non-linear patterns between air pollution, geography and population, the
Random Forest model provided an effective approach to predicting PM10
levels across Europe.


With an RSME of **2.8$\mu g/m^3$**, the model achieved a strong level of predictive accuracy, confirming that this algorithm gives a good generalisation to unseen air quality data. 
Overall, we can confidently say that the Random Forest proved to be a robust and realistic modelling approach for our air quality regression problem. 

